F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-10-25 icfp-2015/paper.tex
Adding alternative XML sample (with variants)

\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage[hang,flushmargin]{footmisc} 
\usepackage{graphics} 
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage[hyphens]{url}
\usepackage{hyperref}
\usepackage{ftnright}
\usepackage[notquote]{hanging}
\usepackage[T1]{fontenc}
\usepackage{semantic}
\usepackage{enumitem}
\usepackage{flushend}

@@

% TODO: The model here does not actually use type erasure!!
% TODO: Explain why we can only provide simple classes (!)
% TODO: Insert a few more mentions of why "s1 + s2 <: s1 + s2 + s3" is valid (cannot do complete pattern match)

@@

@@

\acks
We would like to thank to the F\# Data contributors on GitHub and other colleagues working on type providers,
including Jomo Fisher, Keith Battocchi and Kenji Takeda.

>>>>

\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{graphics} 
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage[hyphens]{url}
\usepackage{hyperref}
\usepackage{ftnright}
\usepackage[T1]{fontenc}
\usepackage{semantic}
\usepackage{enumitem}
%\usepackage[hang,flushmargin]{footmisc} 
%\usepackage[notquote]{hanging}
%\usepackage{flushend}

@@

% TODO: Adapt XML example 
% TODO: Switch from sum types to variant types
% TODO: Discuss 'stability' of the inference (minor change in schema does not cause dramatic cahnges)
% TODO: Say that any variant is a subtype of any other variant 
% TODO: Say that anything is a subtype of variant
% TODO: Say that variant is a last-resort thing
% TODO: Double check that bottom/top are used correctly
% TODO: Explain (even more) that there is no least upper bound, which is why Theorem 3 is good enough
% TODO: Avoid "no other rule applies" in the formal system
% TODO: Practical evaluation (lots of downloads, lots of Github - how often people customize samples?)

@@

\subsection{Processing XML data, alternative}

In XML, documents are formed by (possibly) nested elements with attributes. We can view elements
as records (unlike in JSON, the records are named) with a field for each attribute and an additional
special field for the nested contents (which may be a collection of elements).

As an example, consider a simple extensible document format where a root element
{\ttfamily\small <doc>} can contain a number of document element, one of which is
{\ttfamily\small <heading>} representing headings:
%
{\small{
\begin{verbatim}
  <doc>
    <heading>Working with JSON</heading>
    <p>Type providers make this easy.</p>
    <heading>Working with XML</heading>
    <p>Processing XML is as easy as JSON.</p>
    <image source="xml.png" />
  </doc>
\end{verbatim}
}}
%
\noindent
The F\# Data library has been primarilly designed to allow easy reading of data. For example,
say we want to print all headings that appear in the document. The sample shows a part of the 
possible document structure (in particular the {\ttfamily\small <heading>} element we want to
read), but it does not show all possible elements (say, {\ttfamily\small <table>}). Assuming the
above document is \strf{sample.doc}, we can write:
%
\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Document}~=~\ident{XmlProvider}\langl\str{sample.doc}\rangl\hspace{1em} \\[0.5em]
 \kvd{let}~\ident{root}~=~\ident{Document.Load}(\str{http://.../another.doc})\\
 \kvd{for}~\ident{elem}~\kvd{in}~\ident{root.Doc}~\kvd{do}\\
 \quad\kvd{match}~\ident{elem.Heading}~\kvd{with} \\
 \quad|~\ident{Some text} \rightarrow \ident{printf}~\str{ - \%s }~\ident{text}\\
 \quad|~\ident{None} \rightarrow ()\\
\end{array}
\end{equation*}
%
The example iterates over a collection of elements \ident{elem} returned by \ident{root.Doc}. 
Our system provides a typed access to elements known from the sample document and so we can
write \ident{elem.Heading}.

\paragraph{Open world and variant types.}
In the above example, The system does not infer a union type representing a choice between
heading, paragraph and image. By its nature, the XML format is extensible and the sample cannot 
include all possible examples\footnote{Even when the document structure is defined using XML 
Schema, documents may contain elements prefixed with other namespaces.}. The type of \ident{elem}
is thus a \emph{variant} type annotated with the three statically known possibilities (heading,
paragraph, image). The variant is mapped to the following F\# type:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Element}~=  \\[0.1em]
 \quad \kvd{member}~\ident{Heading}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Paragraph}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Image}~:~\ident{option}\langl \ident{Image} \rangl\\
\end{array}
\end{equation*}
%
The type provides a nice access to the elements known statically from the sample, but it encodes
the \emph{open-world assumption} -- the user also needs to handle the case when the value is none 
of the statically known elements.

One perhaps surprising consequence of the use of variant types is that the above code will work
correctly on document that contains elements other than those in the sample document. This would 
arguably not be desirable design for data types in a programming language, but it is extremely
useful when reading data from external data sources. It is also worth noting that our type 
inference attempts to minimze the number of variant types that appear in the resulting type --
a variant is not used if there is another option.

@@

\acks
?
We would like to thank to the F\# Data contributors on GitHub and other colleagues working on type providers,
including Jomo Fisher, Keith Battocchi and Kenji Takeda.
